\bf Visual diagnostics for constrained optimisation with application to guided tours

### 2020-11-17 paper/paper.Rmd
updating references

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (REF). The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.
<!-- $f: S \rightarrow \mathbb{R}$, with $S \in \mathbb{R}^n$, in a reduced space given by the problem constraints on $f$.  -->

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (REF) which is an exploratory data analysis tool that is defined to detect *interesting structures* or features  in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (REF). The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (REF).

The optimization challenges encounter in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existance of multiple maxima (local and global), the tradeoff between computational burden and proximity to the maxima,  dealing with noisy objective functions that might be non smooth and non diferentiable (REF). Those  are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution to a problem. These numerical methods follow interactive approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existance of multiple maxima (local and global), the trade off between desirable accuracy and computational burden, as well as dealing with noisy objective functions. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

In this paper we focus on unconstrained optimization as defined in the introduction section and assume that it is not possible to find an analytical solution in closed form. That is, our problem consists of finding the minimum of a function $f \in L^p$ in the unconstrained $\mathbb{R}$ space.  In this scenario, gradient descent methods have taken the stage for many decades now. In gradient descent methods the goal is to find the minimum of an objective function, and the way to do so, is by taken steps  proportional to the negative gradient of the function at the evaluation point, that is $x_{k+1} = x_k -\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate (REF). The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. In order to solve this problem line minimization or line search methods were proposed (REF) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$ and then for each iteration a search direction $S_k$ is established then the objective is to find the value of $\alpha_k \in \mathbb{R}$ that minimizes $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{(k)}$, and on approximations of the step size $\alpha_k$, catering for practical optimisation problems. Here we focus on optimisation methods used with the projection pursuit guided tour described in the next section.

Derivative free --> adding

>>>>

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (@kelley1999iterative). The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.
<!-- $f: S \rightarrow \mathbb{R}$, with $S \in \mathbb{R}^n$, in a reduced space given by the problem constraints on $f$.  -->

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (@buja2005computational) which is an exploratory data analysis tool that is defined to detect *interesting structures* or features  in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (cook2008grand). The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (@cook1993projection).

The optimization challenges encounter in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple maxima (local and global), the trade off between computational burden and proximity to the maxima,  dealing with noisy objective functions that might be non smooth and non diferentiable (@jones1998efficient). Those  are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution to a problem. These numerical methods follow interactive approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existence of multiple maxima (local and global), the trade off between desirable accuracy and computational burden, as well as dealing with noisy objective functions. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

In this paper we focus on unconstrained optimization as defined in the introduction section and assume that it is not possible to find a solution in closed form. That is, our problem consists of finding the minimum of a function $f \in L^p$ in the unconstrained $\mathbb{R}$ space.  In this scenario, gradient descent methods have taken the stage for many decades now. In gradient descent methods the goal is to find the minimum of an objective function, and the way to do so, is by taken steps  proportional to the negative gradient of the function at the evaluation point, that is $x_{k+1} = x_k -\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate (@fletcher2013practical). The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned (@trefethen1997numerical). In order to solve some of those problems, line minimization or line search methods were proposed (@shi2004convergence) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that minimizes $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{(k)}$, and on approximations of the step size $\alpha_k$, catering for practical optimisation problems. Here we focus on line search methods for the optimisation used in the projection pursuit guided tour described in the next section.

### 2020-11-17 paper/paper.tex
updating references

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (REF). The unconstrained problem can be formulated as a minimization (or maximization) problem such as
\(min_{x} f(x)\) where \(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with certain properties defined in an \(L^p\) space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions are introduced via a set of functions that can be convex or non-convex: \(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and hence the problem can be written as
\(min_{x} f(x)\) \emph{subject to} \(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (REF) which is an exploratory data analysis tool that is defined to detect \emph{interesting structures} or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (REF). The target of the optimisation is to identify the most \emph{interesting} low-dimensional views of the data given by a corresponding projection matrix. The most \emph{interesting} structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (REF).

The optimization challenges encounter in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existance of multiple maxima (local and global), the tradeoff between computational burden and proximity to the maxima, dealing with noisy objective functions that might be non smooth and non diferentiable (REF). Those are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution to a problem. These numerical methods follow interactive approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existance of multiple maxima (local and global), the trade off between desirable accuracy and computational burden, as well as dealing with noisy objective functions. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

In this paper we focus on unconstrained optimization as defined in the introduction section and assume that it is not possible to find an analytical solution in closed form. That is, our problem consists of finding the minimum of a function \(f \in L^p\) in the unconstrained \(\mathbb{R}\) space. In this scenario, gradient descent methods have taken the stage for many decades now. In gradient descent methods the goal is to find the minimum of an objective function, and the way to do so, is by taken steps proportional to the negative gradient of the function at the evaluation point, that is \(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or the learning rate (REF). The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. In order to solve this problem line minimization or line search methods were proposed (REF) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate \(x_{1}\) and then for each iteration a search direction \(S_k\) is established then the objective is to find the value of \(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) (this step is called the line search). Then move on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\).
Modern development of line search methods focuses on improving the computation of the searching direction \(S_{(k)}\), and on approximations of the step size \(\alpha_k\), catering for practical optimisation problems. Here we focus on optimisation methods used with the projection pursuit guided tour described in the next section.

Derivative free --\textgreater{} adding

>>>>

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (\citet{kelley1999iterative}). The unconstrained problem can be formulated as a minimization (or maximization) problem such as
\(min_{x} f(x)\) where \(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with certain properties defined in an \(L^p\) space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions are introduced via a set of functions that can be convex or non-convex: \(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and hence the problem can be written as
\(min_{x} f(x)\) \emph{subject to} \(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (\citet{buja2005computational}) which is an exploratory data analysis tool that is defined to detect \emph{interesting structures} or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (cook2008grand). The target of the optimisation is to identify the most \emph{interesting} low-dimensional views of the data given by a corresponding projection matrix. The most \emph{interesting} structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (\citet{cook1993projection}).

The optimization challenges encounter in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple maxima (local and global), the trade off between computational burden and proximity to the maxima, dealing with noisy objective functions that might be non smooth and non diferentiable (\citet{jones1998efficient}). Those are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution to a problem. These numerical methods follow interactive approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existence of multiple maxima (local and global), the trade off between desirable accuracy and computational burden, as well as dealing with noisy objective functions. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

In this paper we focus on unconstrained optimization as defined in the introduction section and assume that it is not possible to find a solution in closed form. That is, our problem consists of finding the minimum of a function \(f \in L^p\) in the unconstrained \(\mathbb{R}\) space. In this scenario, gradient descent methods have taken the stage for many decades now. In gradient descent methods the goal is to find the minimum of an objective function, and the way to do so, is by taken steps proportional to the negative gradient of the function at the evaluation point, that is \(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or the learning rate (\citet{fletcher2013practical}). The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned (\citet{trefethen1997numerical}). In order to solve some of those problems, line minimization or line search methods were proposed (\citet{shi2004convergence}) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for each iteration where the objective is to find the value of \(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) (this step is called the line search). Then move on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction \(S_{(k)}\), and on approximations of the step size \(\alpha_k\), catering for practical optimisation problems. Here we focus on line search methods for the optimisation used in the projection pursuit guided tour described in the next section.
